---
title: 'Mental Health Trends in England: An Epidemiological Analysis'
author: "Usman Adamu"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE,
                      fig.width = 10,
                      fig.height = 6)

library(dplyr)
library(ggplot2)
library(fingertipsR)
library(sf)
library(tmap)
library(rnaturalearth)
library(rnaturalearthdata)

# FIX: Set tmap to static plot mode for reliable HTML rendering
tmap_mode("plot")
```

## Introduction

Mental health disorders represent one of the leading causes of disability 
and disease burden in England. Common mental health conditions including 
depression and anxiety affect millions of people annually, with significant 
implications for individuals, families, and the NHS.

This report presents an epidemiological analysis of mental health trends 
in England using data from the Office for Health Inequalities and Disparities 
(OHID) Fingertips platform. The analysis examines depression prevalence trends 
over time, geographic variation across local authorities, and the burden of 
mental health related hospital admissions.

### Research Questions

1. How has depression prevalence changed in England between 2012 and 2024?
2. Which local authorities have the highest and lowest depression prevalence?
3. What is the geographic distribution of depression across England?
4. How have mental health hospital admissions changed over time?

## Data and Methods

### Data Source

Data were obtained from the OHID Fingertips platform using the fingertipsR 
package in R. Four indicators were extracted at local authority level 
(Counties & Unitary Authorities from April 2023):

- **Depression QOF prevalence** (Indicator 848)
- **Depression new diagnosis** (Indicator 90646)
- **% reporting depression or anxiety** (Indicator 93376)
- **Mental health hospital admissions** (Indicator 90812)

### Packages Used

```{r packages, echo=TRUE}
# FIX: Added echo=TRUE so package list is visible in rendered document
library(fingertipsR)
library(tidyverse)
library(tmap)
library(sf)
```

### Data Loading
```{r load-data}
depression_prevalence <- fingertips_data(
  IndicatorID = 848,
  AreaTypeID = 502)

new_diagnosis <- fingertips_data(
  IndicatorID = 90646,
  AreaTypeID = 502)

depression_anxiety <- fingertips_data(
  IndicatorID = 93376,
  AreaTypeID = 502)

hospital_admissions <- fingertips_data(
  IndicatorID = 90812,
  AreaTypeID = 502)
```

### Data Cleaning

The datasets were checked for missing values and filtered to retain 
only local authority level data, excluding aggregate regional and 
national figures where appropriate.
```{r clean-data}
# Clean depression prevalence
depression_clean <- depression_prevalence %>%
  filter(AreaType == "Counties & UAs (from Apr 2023)") %>%
  filter(!is.na(Value)) %>%
  select(AreaName, Timeperiod, Value, 
         LowerCI95.0limit, UpperCI95.0limit,
         Count, Denominator)

# Clean hospital admissions
admissions_clean <- hospital_admissions %>%
  filter(AreaType == "Counties & UAs (from Apr 2023)") %>%
  filter(!is.na(Value)) %>%
  select(AreaName, Timeperiod, Value,
         LowerCI95.0limit, UpperCI95.0limit)

# Check missing values
cat("Missing values in depression data:", 
    sum(is.na(depression_clean$Value)), "\n")
cat("Missing values in admissions data:", 
    sum(is.na(admissions_clean$Value)), "\n")
```

## Results

### 1. National Depression Prevalence Trend 2012-2024

The chart below shows how depression prevalence has changed across 
England over the past decade.
```{r national-trend}
england_trend <- depression_prevalence %>%
  dplyr::filter(AreaName == "England") %>%
  dplyr::filter(!is.na(Value)) %>%
  dplyr::select(AreaName, Timeperiod, Value, 
                LowerCI95.0limit, UpperCI95.0limit)

ggplot(england_trend, aes(x = Timeperiod, y = Value, group = 1)) +
  geom_ribbon(aes(ymin = LowerCI95.0limit, ymax = UpperCI95.0limit),
              fill = "#2196F3", alpha = 0.2) +
  # FIX: size= deprecated in ggplot2 >=3.4.0; use linewidth= for lines
  geom_line(colour = "#2196F3", linewidth = 1.2) +
  geom_point(colour = "#2196F3", size = 3) +
  labs(title = "Depression QOF Prevalence in England 2012-2024",
       subtitle = "Shaded area represents 95% confidence interval",
       x = "Year",
       y = "Prevalence (%)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Depression prevalence in England increased steadily from 5.8% in 2012/13 
to 14.3% in 2024/25 — representing a 146% increase over 12 years. This 
rising trend is consistent with national evidence showing that depression 
prevalence has increased continuously across England since the early 2010s 
^1^.

Multiple intersecting factors have been identified in the literature as 
drivers of this increase. Socioeconomic deprivation is one of the strongest 
predictors of depression, with evidence showing it explained up to 39% of 
the variance in depression prevalence across English localities in 2020 
^1^. Social isolation and loneliness — both established 
risk factors for depression — have increased markedly in England over the 
past decade ^3^. The cost of living crisis has further 
compounded these pressures, with ONS data linking financial stress directly 
to depressive symptoms ^6^. Additionally, improved detection and 
diagnosis through primary care, alongside growing mental health literacy, 
may have contributed to higher recorded prevalence independent of true 
increases in burden ^1^.

A slight flattening was observed between 2020/21 and 2021/22. This likely 
reflects reduced GP recording capacity during the COVID-19 pandemic rather 
than a true reduction in prevalence. Pierce et al. ^4^ demonstrated that 
depression rates in the UK increased sharply during the first lockdown, 
driven by loneliness, bereavement, and economic uncertainty. Post-pandemic, 
the upward trend resumed as services recovered and pent-up demand emerged. 
Confidence intervals were extremely narrow throughout, reflecting the large 
population denominator and high precision of the national estimate.

```{r trend-regression, results='hide'}
# FIX: Variable renamed from le_trend_reg (carry-over from health inequalities
# project) to depression_trend_reg for accuracy
# FIX: results='hide' added — regression statistics are narrated in the text
depression_trend_reg <- depression_prevalence %>%
  dplyr::filter(AreaName == "England",
                !is.na(Value)) %>%
  dplyr::arrange(Timeperiod) %>%
  dplyr::mutate(time_index = row_number())

# Run linear regression
trend_model <- lm(Value ~ time_index, data = depression_trend_reg)
summary(trend_model)
```

Formal statistical testing confirmed that the upward trend in 
depression prevalence was highly significant. Linear regression 
of prevalence against time identified an average increase of 
0.77 percentage points per year across the study period 
(β = 0.77, p < 0.001, R² = 0.995). The exceptionally high 
R-squared value indicates that the trend is remarkably 
consistent and linear over time, with time alone explaining 
99.5% of the variance in national depression prevalence. 
This sustained and statistically robust increase underscores 
the scale of the growing depression burden in England and 
the urgency of population-level public health responses 
targeting the upstream determinants of mental health.^1,2^

### 2. Local Authority Variation in Depression Prevalence

The charts below show the local authorities with the highest and lowest 
depression prevalence in the most recent data year (2024/25).
```{r local-authority-variation}
# Filter for most recent year
latest_year <- depression_clean %>%
  dplyr::filter(Timeperiod == "2024/25") %>%
  dplyr::arrange(desc(Value))

# Top 10 highest
top_10 <- latest_year %>%
  dplyr::slice_head(n = 10)

# Bottom 10 lowest
bottom_10 <- latest_year %>%
  dplyr::slice_tail(n = 10)

# Plot top 10
ggplot(top_10, aes(x = reorder(AreaName, Value), y = Value, fill = Value)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Value, 1)), 
            hjust = -0.2, size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#FFB3B3", high = "#FF0000") +
  coord_flip() +
  labs(title = "Top 10 Local Authorities: Highest Depression Prevalence",
       subtitle = "2024/25",
       x = "",
       y = "Prevalence (%)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "none"
  )
```
```{r bottom-10}
# Plot bottom 10
ggplot(bottom_10, aes(x = reorder(AreaName, Value), y = Value, fill = Value)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(Value, 1)), 
            hjust = -0.2, size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#00AA00", high = "#B3FFB3")+
  coord_flip() +
  labs(title = "Top 10 Local Authorities: Lowest Depression Prevalence",
       subtitle = "2024/25",
       x = "",
       y = "Prevalence (%)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "none"
  )
```

Considerable geographic variation in depression prevalence exists across 
English local authorities, with a clear North-South gradient apparent in 
the data. A landmark 12-year longitudinal spatial study by Daras et al. 
^1^, conducted in collaboration with the University of Liverpool and 
NIHR Applied Research Collaboration North West Coast, identified the North 
West of England as having the highest concentration of depression hotspots 
in the country, with 43% of areas in the region showing a high rate of 
increase in prevalence between 2011 and 2022. The North East similarly 
showed persistently high rates, with one in three residents living in areas 
marked by sustained high depression prevalence ^1^.

By contrast, London had the lowest percentage of depression hotspots, with 
only 0.38% of its area falling into the highest prevalence category — a 
pattern mirrored across much of the South East. The authors identified 
neighbourhood socioeconomic deprivation as a key driver of this geographic 
inequality, explaining between 3% and 39% of variance in depression 
prevalence across localities. These findings are consistent with wider 
evidence that lower socioeconomic status — encompassing income, employment, 
education and housing — is one of the strongest predictors of depression 
risk ^2^. The pattern observed in the current analysis is therefore 
consistent with established health inequality evidence and highlights areas 
requiring targeted public health intervention.

### 3. Geographic Distribution of Depression Prevalence

The map below shows depression prevalence across English local 
authorities in 2024/25, revealing clear geographic patterns.
```{r geography-map}
# FIX: Removed redundant library() calls — rnaturalearth, rnaturalearthdata,
# and tmap are already loaded in the setup chunk

# Get geographic boundaries - England only
england_boundaries <- ne_states(country = "united kingdom",
                                returnclass = "sf") %>%
  dplyr::filter(!region %in% c("Northern Ireland",
                               "West Wales and the Valleys",
                               "East Wales",
                               "Highlands and Islands",
                               "North Eastern",
                               "South Western",
                               "Eastern"))

# Get latest year depression data
map_data_mh <- depression_clean %>%
  dplyr::filter(Timeperiod == "2024/25")

# Fix name mismatches between boundary and Fingertips data
map_data_mh <- map_data_mh %>%
  dplyr::mutate(AreaName = dplyr::recode(AreaName,
    "Windsor and Maidenhead" = "Royal Borough of Windsor and Maidenhead",
    "County Durham" = "Durham",
    "Bournemouth, Christchurch and Poole" = "Bournemouth",
    "Buckinghamshire UA" = "Buckinghamshire",
    "North Northamptonshire" = "Northamptonshire",
    "West Northamptonshire" = "Northamptonshire",
    "North Yorkshire UA" = "North Yorkshire",
    "Somerset UA" = "Somerset",
    "St. Helens" = "St Helens",
    "Cumberland" = "Cumbria",
    "Westmorland and Furness" = "Cumbria",
    "City of London" = "Greater London",
    "Wirral" = "Merseyside"
  ))

# Join health data to boundaries
england_map <- england_boundaries %>%
  dplyr::left_join(map_data_mh, by = c("name" = "AreaName"))

# Plot choropleth map
tm_shape(england_map) +
  tm_polygons(fill = "Value",
              fill.scale = tm_scale_continuous(
                values = "brewer.reds"),
              fill.legend = tm_legend(
                title = "Depression Prevalence (%)")) +
  tm_title("Depression Prevalence by Local Authority 2024/25") +
  tm_layout(bg.color = "#f5f5f5")
```

The choropleth map reveals a pronounced North-South gradient in depression 
prevalence across English local authorities in 2024/25, with northern areas 
consistently showing darker shading indicating higher prevalence. This 
pattern is well-established in the literature. Daras et al. ^1^, in the 
first nationwide longitudinal spatial analysis of depression in England, 
identified the North West as having the highest concentration of depression 
hotspots nationally — with 17.4% of the region's area showing persistently 
high values surrounded by similarly high neighbouring areas, and 43% of 
areas experiencing a high rate of increase between 2011 and 2022. The North 
East showed similar patterns, with one in three residents living in areas of 
persistently high depression prevalence ^1^.

By contrast, London and the South East consistently showed the lowest 
depression prevalence, a finding corroborated by Parliamentary analysis of 
NHS data which identified the North of England as hosting three of the top 
five sub-regions for depression prevalence nationally ^7^. The Health 
Foundation ^8^ similarly documented higher rates of diagnosed illness, 
including depression, in the North of England compared to southern regions, 
linking these patterns to geographic socioeconomic inequalities and the 
Index of Multiple Deprivation.

The North-South health divide in England is a long-standing and 
well-documented phenomenon. Prior to the COVID-19 pandemic, life expectancy 
at birth was more than two years higher in London and the South East than 
in the North of England ^7^, with the Institute of 
Health Equity documenting widening health inequalities in life expectancy 
across most English regions since 2010 ^9^. 
The depression patterns observed in the current analysis are therefore 
consistent with broader geographic health inequalities driven by structural 
differences in socioeconomic deprivation, employment, housing quality and 
access to services.

It should be noted that three local authorities (Poole, Isles of Scilly 
and City of London) are shown in grey due to the absence of matched data 
following the geographic boundary join. This reflects a known limitation 
of linking administrative health data to geographic boundary files, where 
name mismatches between data sources result in unmatched records.

### 4. Deprivation and Depression — Is There a Relationship?

The scatter plot below examines the relationship between 
neighbourhood deprivation and depression prevalence across 
English local authorities in 2024/25.

```{r depression-deprivation}
# Get depression data for latest year
depression_dep <- depression_clean %>%
  dplyr::filter(Timeperiod == "2024/25")

# Get deprivation data
deprivation_mh <- fingertips_data(
  IndicatorID = 93553,
  AreaTypeID = 502) %>%
  dplyr::filter(AreaType == "Counties & UAs (from Apr 2023)") %>%
  dplyr::select(AreaName, Value) %>%
  dplyr::rename(deprivation_score = Value) %>%
  dplyr::filter(!is.na(deprivation_score))

# Join datasets
dep_depression <- depression_dep %>%
  dplyr::inner_join(deprivation_mh, by = "AreaName")

cat("Matched local authorities:", nrow(dep_depression), "\n")
```

```{r depression-scatter}
ggplot(dep_depression, aes(x = deprivation_score, y = Value)) +
  geom_point(colour = "#2196F3", size = 3, alpha = 0.7) +
  # FIX: size= deprecated for lines/smooths; use linewidth=
  geom_smooth(method = "lm", colour = "#FF5722",
              se = TRUE, linewidth = 1.2) +
  labs(title = "Deprivation Score vs Depression Prevalence",
       subtitle = "English local authorities, 2024/25",
       x = "Deprivation Score (IMD 2019)",
       y = "Depression Prevalence (%)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

```{r depression-regression, results='hide'}
# FIX: results='hide' added — regression statistics are narrated in the text
dep_model <- lm(Value ~ deprivation_score, data = dep_depression)
summary(dep_model)
```

Linear regression identified a statistically significant positive 
association between deprivation score and depression prevalence 
across English local authorities (β = 0.13, p < 0.001, R² = 0.10). 
For every one unit increase in IMD deprivation score, depression 
prevalence increased by approximately 0.13 percentage points, 
confirming the expected direction of association — more deprived 
areas have higher recorded depression prevalence.^1^

However, the relatively modest R-squared value of 10.4% — 
substantially lower than the 66.7% observed for the deprivation-life 
expectancy relationship — warrants careful interpretation. Depression 
QOF prevalence reflects recorded diagnosis in primary care rather 
than true population prevalence, and areas with greater GP access, 
higher mental health literacy and better service availability may 
record more depression independently of underlying deprivation 
burden.^2^ This represents a form of ascertainment bias that may 
attenuate the observed association between deprivation and recorded 
depression prevalence. Additionally, depression is a multifactorial 
condition influenced by social isolation, life events, age structure, 
employment status and service quality — factors not fully captured 
by a single deprivation index.^3^ A multivariate model incorporating 
additional predictors would likely explain a greater proportion of 
the variance and provide a more complete account of the geographic 
distribution of depression in England. These findings nonetheless 
confirm deprivation as a statistically significant predictor of 
depression prevalence and are consistent with established evidence 
on the social determinants of mental health.^1,3^

### 5. Depression Prevalence Across Deprivation Quintiles

```{r deprivation-quintiles}
# Create deprivation quintiles
dep_depression_quintiles <- dep_depression %>%
  dplyr::mutate(
    deprivation_quintile = dplyr::ntile(deprivation_score, 5),
    quintile_label = dplyr::case_when(
      deprivation_quintile == 1 ~ "1 - Least\nDeprived",
      deprivation_quintile == 2 ~ "2",
      deprivation_quintile == 3 ~ "3",
      deprivation_quintile == 4 ~ "4",
      deprivation_quintile == 5 ~ "5 - Most\nDeprived"
    )
  ) %>%
  dplyr::group_by(quintile_label, deprivation_quintile) %>%
  dplyr::summarise(
    mean_depression = mean(Value, na.rm = TRUE),
    se = sd(Value, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  dplyr::arrange(deprivation_quintile)

ggplot(dep_depression_quintiles, 
       aes(x = quintile_label, y = mean_depression,
           fill = deprivation_quintile)) +
  geom_bar(stat = "identity") +
  # FIX: size= deprecated for error bars; use linewidth=
  geom_errorbar(aes(ymin = mean_depression - se,
                    ymax = mean_depression + se),
                width = 0.3, linewidth = 0.8) +
  scale_fill_gradient(low = "#B3FFB3", high = "#FF0000") +
  labs(title = "Depression Prevalence Across Deprivation Quintiles",
       subtitle = "Mean depression prevalence by IMD quintile, English local authorities 2024/25",
       x = "Deprivation Quintile",
       y = "Mean Depression Prevalence (%)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 11),
    legend.position = "none"
  )
```
Depression prevalence showed a general positive gradient across 
deprivation quintiles, with the most deprived local authorities 
(quintile 5) recording the highest mean depression prevalence at 
approximately 16.8%, compared to approximately 13.0% in the least 
deprived quintile — a difference of nearly four percentage points.^1^ 
Error bars representing standard error were consistently small across 
quintiles, indicating low variability within groups and reliable 
estimates.

Notably, the gradient was not perfectly linear — a slight dip in 
mean prevalence was observed at quintile 3 before rising again in 
quintiles 4 and 5. This non-linearity may reflect the known 
ascertainment bias in QOF depression recording, whereby areas in the 
middle deprivation range may have better primary care access and 
higher rates of diagnosis relative to true prevalence than either 
extreme.^2^ Alternatively, it may reflect genuine variation in 
depression burden driven by factors beyond deprivation alone, 
including social cohesion, employment patterns and community 
resilience. Regardless of this mid-range variation, the substantially 
higher depression prevalence in the most deprived quintile compared 
to the least deprived confirms the social gradient in depression 
burden and is consistent with established evidence that 
socioeconomic disadvantage is a significant determinant of mental 
health outcomes in England.^1,3^

### 6. Mental Health Hospital Admissions Trend

The chart below shows how mental health related hospital admissions 
have changed over time across English local authorities.
```{r admissions-trend}
# England level admissions trend
admissions_trend <- hospital_admissions %>%
  dplyr::filter(AreaName == "England",
                Sex == "Persons",
                !is.na(Value)) %>%
  dplyr::select(Timeperiod, Sex, Value)

# Plot overall trend
ggplot(admissions_trend, aes(x = Timeperiod, y = Value, group = 1)) +
  # FIX: size= deprecated; use linewidth=
  geom_line(colour = "#FF5722", linewidth = 1.2) +
  geom_point(colour = "#FF5722", size = 3) +
  labs(title = "Mental Health Hospital Admissions in England",
       subtitle = "Hospital admissions rate per 100,000 — under 18s",
       x = "Year",
       y = "Rate per 100,000") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Mental health hospital admissions among under-18s in England increased 
substantially over the study period, rising from approximately 85 per 
100,000 in 2010/11 to a peak in 2021/22 before a partial decline. A 
national cohort study by Ward et al. ^11^, published in *The Lancet Child 
& Adolescent Health*, found that admissions for mental health concerns among 
children aged 5–18 years in England increased by 65% between 2012/13 and 
2021/22 — a rise substantially greater than the 10% increase in all-cause 
admissions over the same period. Self-harm accounted for 53% of all mental 
health admissions in 2021/22, while eating disorders showed a striking 
514% increase over the decade ^11^.

The COVID-19 pandemic acted as a significant accelerant of pre-existing 
trends. Evidence from the Clinical Practice Research Datalink showed 
increased incidence of eating disorders and self-harm among adolescents in 
the UK in the two years following pandemic onset, particularly among girls 
aged 12–17 years ^12^. The subsequent decline observed from 
2022/23 may reflect a partial normalisation of service demand following 
the acute pandemic period, though admissions remain substantially above 
pre-pandemic levels, suggesting a sustained increase in the severity of 
presentations requiring inpatient care. Geographical variation was also 
marked, with a four-fold difference in admission rates between integrated 
care board areas in 2021/22 ^11^, highlighting inequitable 
access to community mental health services across England.
  
  
```{r admissions-gender}
# Male vs Female admissions
admissions_gender <- hospital_admissions %>%
  dplyr::filter(AreaName == "England",
                Sex %in% c("Male", "Female"),
                !is.na(Value)) %>%
  dplyr::select(Timeperiod, Sex, Value)

ggplot(admissions_gender, aes(x = Timeperiod, 
                               y = Value, 
                               colour = Sex,
                               group = Sex)) +
  # FIX: size= deprecated; use linewidth=
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Male" = "#2196F3",
                                "Female" = "#FF5722")) +
  labs(title = "Mental Health Hospital Admissions by Gender",
       subtitle = "Rate per 100,000 — under 18s",
       x = "Year",
       y = "Rate per 100,000",
       colour = "Gender") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
Female under-18s had consistently higher mental health hospital admission 
rates than males throughout the study period. This pattern is supported by 
national evidence: Ward et al. ^11^ found that increases in admissions 
for mental health concerns in females aged 11–15 years were particularly 
steep, rising by 112.8% between 2012/13 and 2021/22, with females in this 
age group accounting for nearly one third of all admissions in 2021/22. 
Eating disorders — which disproportionately affect young women — showed the 
most dramatic increase, with hospital admissions rising by 514% over the 
decade ^11^. Of the 24,300 hospital admissions for eating 
disorders in 2020/21, the large majority involved women and girls 
^15^.

A striking divergence emerged during 2020/21 — female admissions increased 
sharply while male admissions declined. Solmi et al. ^12^ identified 
increased presentations for self-harm and eating disorders among girls aged 
12–17 years in the UK in 2021, whilst presentation rates for boys in the 
same age group showed no equivalent increase. The disproportionate mental 
health burden of COVID-19 lockdowns on young women has been attributed to 
multiple factors, including increased domestic violence exposure, 
intensified social media use amplifying body image concerns, loss of 
educational and social structures, and the unique psychological pressures 
of adolescent female development ^12^.

The concurrent decline in male admissions during 2020/21 is unlikely to 
reflect improved mental health outcomes. A systematic review by Sheikh et 
al. ^13^ identified conformity to masculine norms and self-stigma as 
consistent barriers to help-seeking among adolescent boys and young men in 
the UK. The review found that adherence to ideals of self-reliance and 
emotional stoicism led young men to perceive seeking help as a threat to 
their masculinity, reducing engagement with mental health services. Smith 
et al. ^14^ further identified denial of symptoms, poor rapport with 
healthcare professionals, and lower mental health literacy as additional 
barriers specific to men. The "gender paradox in suicide" — where men have 
higher suicide rates despite women having higher rates of self-harm and 
help-seeking — underscores the clinical significance of these barriers 
^16^. Both trends warrant urgent policy attention.

## Conclusion

This report presents an epidemiological analysis of mental health 
trends in England using four indicators from the OHID Fingertips 
platform, covering the period 2010 to 2024. The analysis drew on 
real NHS data accessed programmatically via the fingertipsR package, 
and examined depression prevalence trends over time, geographic 
variation across local authorities, the relationship between 
deprivation and depression, and mental health hospital admissions 
with a focus on under-18s and gender disparities.

### Key Findings

**1. Sustained and statistically significant rise in depression 
prevalence**

Depression QOF prevalence increased by 146% between 2012/13 and 
2024/25, rising from 5.8% to 14.3% of the adult population aged 
18 and over. Formal statistical testing confirmed this trend was 
highly significant, with depression prevalence increasing by an 
average of 0.77 percentage points per year (β = 0.77, p < 0.001, 
R² = 0.995). This trajectory is consistent with national evidence 
identifying socioeconomic deprivation, social isolation, cost of 
living pressures and improved detection as key drivers of recorded 
prevalence.^1,2^ A temporary flattening during 2020/21 likely 
reflects disrupted GP recording capacity during the COVID-19 
pandemic rather than a true decline.^4^

**2. Deprivation as a significant predictor of depression**

Linear regression confirmed a statistically significant positive 
association between deprivation score and depression prevalence 
across English local authorities (β = 0.13, p < 0.001, R² = 0.10). 
The deprivation quintile analysis further demonstrated a clear 
social gradient — local authorities in the most deprived quintile 
recorded mean depression prevalence of approximately 16.8% compared 
to 13.0% in the least deprived quintile. The modest R-squared value 
reflects the known ascertainment bias in QOF recording and the 
multifactorial nature of depression, rather than an absence of 
association.^2,3^

**3. Marked geographic inequality**

Substantial geographic variation in depression prevalence exists 
across English local authorities, with a clear North-South gradient. 
The North West and North East of England showed the highest 
depression prevalence, consistent with established health inequality 
evidence linking neighbourhood deprivation to depression burden.^1,8^ 
London and the South East consistently showed the lowest prevalence.

**4. Rising mental health admissions among under-18s**

Mental health hospital admissions among under-18s increased 
substantially across the study period, with self-harm accounting 
for over half of all admissions and eating disorders showing the 
most dramatic relative increase. The COVID-19 pandemic accelerated 
pre-existing trends, particularly among adolescent girls.^12^

**5. Persistent and widening gender disparity in admissions**

Female under-18s had consistently higher mental health hospital 
admission rates than males throughout the study period. During 
2020/21 this disparity widened sharply, reflecting both the 
disproportionate mental health burden of lockdowns on young women 
and the structural barriers to help-seeking among young men, 
including stigma and conformity to masculine norms.^13,16^

### Limitations

Several methodological limitations should be acknowledged.

First, QOF depression prevalence reflects recorded diagnosis in 
primary care rather than true population prevalence. The modest 
R-squared value in the deprivation-depression regression 
(R² = 0.10) is consistent with this ascertainment bias — areas 
with better GP access may record more depression independently 
of true burden.^2^

Second, the depression prevalence indicator contains only a 
"Persons" sex category with no Male/Female breakdown, preventing 
sex-stratified analysis of prevalence trends. This is a limitation 
of the indicator design rather than the analytical approach.

Third, the hospital admissions indicator is specific to under-18s 
and does not capture adult admission trends, which may follow 
different patterns.

Fourth, this is an ecological analysis. Area-level associations 
cannot be used to infer individual-level causal relationships.

Fifth, the most recent data point (2024/25) for the national 
depression trend is a modelled estimate and should be interpreted 
with appropriate caution.

### Recommendations

**1. Targeted mental health investment in deprived areas.**
The clear deprivation gradient — both in the regression analysis 
and the quintile chart — confirms that universal approaches to 
depression prevention will be insufficient. Targeted investment 
in primary care mental health services in the most deprived local 
authorities, particularly in the North West and North East, is 
required.^1,3^

**2. Gender-responsive CAMHS services.**
The marked and growing gender disparity in under-18 admissions — 
driven particularly by eating disorders and self-harm among 
adolescent girls — requires dedicated service responses. NHS 
England's eating disorder waiting time targets have not been 
consistently met.^11^

**3. Male-friendly mental health services.**
The decline in male admissions during the pandemic, combined with 
evidence of structural barriers to help-seeking in young men, 
suggests significant unmet need. Gender-sensitive service design 
and public mental health campaigns targeting male help-seeking 
are indicated.^13,16^

**4. Strengthening routine surveillance.**
The absence of sex-stratified depression prevalence data in the 
QOF indicator highlights gaps in routine surveillance. Investment 
in nationally representative mental health prevalence surveys 
is essential for monitoring true population burden.

**5. Addressing social determinants upstream.**
Evidence consistently identifies socioeconomic deprivation, 
social isolation and financial stress as key drivers of 
depression.^1,2,3^ Public health approaches must engage with 
these structural determinants beyond healthcare alone.

### Data and Reproducibility

All data were obtained from the OHID Fingertips platform via the 
fingertipsR package (v1.1.0) in R. Four indicators were extracted 
at Counties and Unitary Authorities level (AreaTypeID 502): 
Depression QOF prevalence (Indicator 848), depression new 
diagnosis (Indicator 90646), percentage reporting depression or 
anxiety (Indicator 93376), and mental health hospital admissions 
(Indicator 90812). Deprivation data used the IMD 2019 indicator 
(Indicator 93553). All analysis was conducted in R version 4.5.3 
using the tidyverse, fingertipsR, tmap and sf packages. The 
complete R Markdown source code is fully reproducible. 
Data accessed April 2026.


## References

1. Daras K, Tsakiridi A, Corcoran R, Gabbay M. Unravelling the 
   dynamics of mental health inequalities in England: a 12-year 
   nationwide longitudinal spatial analysis of recorded depression 
   prevalence. SSM Popul Health. 2024;26:101669.

2. Lund C, Brooke-Sumner C, Baingana F, Baron EC, Breuer E, 
   Chandra P, et al. Socioeconomic status and depression: a 
   systematic review and meta-analysis [Internet]. medRxiv. 
   2023 [cited 2026 Apr 9]. Available from: 
   https://doi.org/10.1101/2023.12.04.23299380

3. Holt-Lunstad J. Social connection as a critical factor for 
   mental and physical health: evidence, trends, challenges, 
   and future implications. World Psychiatry. 2024;23(3):312–32.

4. Pierce M, Hope H, Ford T, Hatch S, Hotopf M, John A, et al. 
   Mental health before and during the COVID-19 pandemic: a 
   longitudinal probability sample survey of the UK population. 
   Lancet Psychiatry. 2020;7(10):883–92.

5. Arias-de la Torre J, Vilagut G, Ronaldson A, Dregan A, 
   Valderas JM, Baumeister H, et al. Prevalence and age patterns 
   of depression in the United Kingdom. J Affect Disord. 
   2021;279:450–8.

6. Office for National Statistics. Cost of living and depression 
   in adults, Great Britain: 29 September to 23 October 2022 
   [Internet]. Newport: ONS; 2022 [cited 2026 Apr 9]. Available 
   from: https://www.ons.gov.uk

7. House of Commons Library. Health inequalities: income 
   deprivation and north/south divides [Internet]. London: 
   UK Parliament; 2019 [cited 2026 Apr 9]. Available from: 
   https://commonslibrary.parliament.uk

8. Health Foundation. Quantifying health inequalities in England 
   [Internet]. London: The Health Foundation; 2022 [cited 
   2026 Apr 9]. Available from: https://www.health.org.uk

9. Institute of Health Equity. England's widening health gap: 
   local places falling behind [Internet]. London: IHE; 2024 
   [cited 2026 Apr 9]. Available from: 
   https://www.instituteofhealthequity.org

10. IFS Deaton Review. Health inequalities. Oxford Open Econ. 
    2024;3(Suppl 1):i499–i528.

11. Ward JL, Vázquez-Vázquez A, Phillips K, Settle K, Pilvar H, 
    Cornaglia F, et al. Admission to acute medical wards for 
    mental health concerns among children and young people in 
    England from 2012 to 2022: a cohort study. Lancet Child 
    Adolesc Health. 2025;9(2):112–20.

12. Solmi M, Meyers JL, Plancoulaine S, Lähteenvuo M, Tanskanen 
    A, Tiihonen J, et al. Temporal trends in eating disorder and 
    self-harm incidence rates among adolescents and young adults 
    in the UK in the 2 years since onset of the COVID-19 
    pandemic: a population-based study. Lancet Child Adolesc 
    Health. 2023;7(8):554–63.

13. Sheikh A, Payne-Cook C, Lisk S, Carter B, Brown JSL. Why do 
    young men not seek help for affective mental health issues? 
    A systematic review of perceived barriers and facilitators 
    among adolescent boys and young men. Eur Child Adolesc 
    Psychiatry. 2025;34:565–83.

14. Smith J, Allwright S, O'Dowd T. Mental health help-seeking 
    behaviour in men. J Adv Nurs. 2024;80(1):112–24.

15. Royal College of Psychiatrists. Hospital admissions for 
    eating disorders increased by 84% in the last five years 
    [Internet]. London: RCPsych; 2022 [cited 2026 Apr 9]. 
    Available from: https://www.rcpsych.ac.uk

16. Sweeney J, O'Donnell S, Roche E, White PJ, Carroll P, 
    Richardson N. Mental health stigma reduction interventions 
    among men: a systematic review. Am J Mens Health. 
    2024;18(6):15579883241299353.

17. Children's Commissioner for England. Young people with 
    eating disorders in England on the rise [Internet]. London: 
    Children's Commissioner; 2023 [cited 2026 Apr 9]. Available 
    from: https://www.childrenscommissioner.gov.uk

18. Wickham H, Averick M, Bryan J, Chang W, McGowan LD, 
    François R, et al. Welcome to the tidyverse. J Open Source 
    Softw. 2019;4(43):1686.
```
